Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-01 Manuscript/03-Simulation.Rmd
ENAR Submission Version

In this section, we compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for nonlinear effect interpolation and prediction. SB-GAM is the first spike-and-slab lasso GAM. We implemented COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence used the default choices. Both mgcv and proposed models allow customized smoothing functions and we chose the cubic regression spline. We controlled the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. 5-fold cross-validation were used for COSSO models, SB-GAM and the proposed models for tuning parameter selection based on the default selection criteria. 20 default candidates of tuning parameters ($s_0$ in BHAM, $\lambda_0$ in SB-GAM) were examined for SB-GAM and the proposed models that allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].
<!-- Other related methods for high-dimensional GAMs also exist, notably the methods of sparse additive models by Ravikumar et al. [@Ravikumar2009], stochastic search term selection for GAM [@Scheipl2012]. However, we exclude these methods from current simulation study because of demonstrated inferior predictive performance compared to mgcv and scalability issues with increased number of predictors. [@Scheipl2013] -->

<!-- ## Monte Carlo Simulation Study -->
We follow the data generating process described in Bai [@Bai2021]. We first generated $n=500$ training data points with $p=4, 10, 50, 200$ predictors respectively, where the predictors $X$ are simulated from a multivariate normal distribution $\text{MVN}_{n\times p}(0, I_{P})$. We then simulated the outcome $y$ from two distributions, Gaussian and binomial with the identity link and logit link $g(x) = \log(\frac{x}{1-x})$ respectively. The mean of each outcome were simulated via the following function
$$
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
$$
for gaussian and binomial outcomes.
<!-- and shrinked effects -->
<!-- $$ -->
<!-- \mathbb{E}(Y) = g^{-1}(\sin(2\pi x_1) - 2\cos(2\pi x_2 -0.5) + 0.4(x_3-0.5) - 2(x_4^2 -0.3)) -->
<!-- $$ -->
<!-- for  poisson outcomes.  -->
Gaussian outcomes required specification of dispersion, where we set the dispersion parameter to be 1. In this data generating process, we have $x_1, x_2, x_3, x_4$ as the active covariates, while the rest covariates are inactive, i.e. $f_j(x_j) = 0$ for $j = 4, \dots, p$. Another set of independent sample of size $n_{test}=1000$,($x_{new} ,y_{new}$), are created following the same data generating process, serving as the testing data. We generated 50 independent pairs of training and testing datasets to evaluate the prediction performance of the chosen models, where training datasets are used to fit the models and testing datasets used to calculate assessment measures.
<!-- The dimensional of the smoothing functions are controlled as k=10. Hence, the corresponding number of coefficient parameters to be estimated is 37, 82, 451 and 1801.  -->
<!-- We also run the simulation where the number of bases are increased to 25 deliberately. -->

@@

To evaluate the  predictive performance of the models, the statistics caluclated based on the testing dataset, $R^2$ for Gaussian model, area under the curve (AUC) for binomial model,  are calculated <!--and deviance are calculated for Poisson model,--> averaging across 50 simulations. The results are tabulated in Table \ref{tab:gaus}, \ref{tab:bin_auc}.

@@

The predictive performances have a consistent pattern across the three distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined. To note, mgcv doesn't support high-dimensional analysis when the number of coefficients are greater than the sample size. mgcv predicts well when $p$ is small or moderate (p = 4, 10), and deteriorate when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are small, moderate, and high. However, in the hyper high dimensional case, SB-GAM has better performance. Among the proposed methods, BHAM-IWLS performs consistently better than BHAM-CD.

>>>>

In this section, we conduct Monte Carlo simulation to compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for nonlinear effect interpolation and prediction. SB-GAM is the first spike-and-slab lasso GAM. We implemented COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence used the default choices. Both mgcv and proposed models allow customized smoothing functions and we chose the cubic regression spline in this study. We controlled the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. 5-fold cross-validation were used for COSSO models, SB-GAM and the proposed models for tuning parameter selection based on the default selection criteria. 20 default candidates of tuning parameters ($s_0$ in BHAM, $\lambda_0$ in SB-GAM) were examined for SB-GAM and the proposed models that allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].
<!-- Other related methods for high-dimensional GAMs also exist, notably the methods of sparse additive models by Ravikumar et al. [@Ravikumar2009], stochastic search term selection for GAM [@Scheipl2012]. However, we exclude these methods from current simulation study because of demonstrated inferior predictive performance compared to mgcv and scalability issues with increased number of predictors. [@Scheipl2013] -->

<!-- ## Monte Carlo Simulation Study -->
We follow the data generating process described in Bai [@Bai2021]. We first generat $n=500$ training data points with $p=4, 10, 50, 200$ predictors respectively, where the predictors $\bs X$ are simulated from a multivariate normal distribution $\text{MVN}_{n\times p}(0, I_{P})$. We then simulat the outcome $y$ from two distributions, Gaussian and binomial with the identity link and logit link $g(x) = \log(\frac{x}{1-x})$ respectively. The mean of each outcome were simulated via the following function
$$
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3)).
$$
<!-- for gaussian and binomial outcomes. -->
<!-- and shrinked effects -->
<!-- $$ -->
<!-- \mathbb{E}(Y) = g^{-1}(\sin(2\pi x_1) - 2\cos(2\pi x_2 -0.5) + 0.4(x_3-0.5) - 2(x_4^2 -0.3)) -->
<!-- $$ -->
<!-- for  poisson outcomes.  -->
Gaussian outcomes required specification of dispersion, where we set the dispersion parameter to be 1. In this data generating process, we have $x_1, x_2, x_3, x_4$ as the active predictors, while the rest predictors are inactive, i.e. $B_j(x_j) = 0$ for $j = 4, \dots, p$. Another set of independent sample of size $n_{test}=1000$,($x_{new} ,y_{new}$), are created following the same data generating process, serving as the testing data. We generated 50 independent pairs of training and testing datasets to evaluate the prediction performance of the chosen models, where training datasets are used to fit and tune the models and testing datasets are used to calculate assessment measures.
<!-- The dimensional of the smoothing functions are controlled as k=10. Hence, the corresponding number of coefficient parameters to be estimated is 37, 82, 451 and 1801.  -->
<!-- We also run the simulation where the number of bases are increased to 25 deliberately. -->

To evaluate the  predictive performance of the models, the statistics calculated out of the testing datasets, $R^2$ for Gaussian model, area under the curve (AUC) for binomial model,  are calculated <!--and deviance are calculated for Poisson model,--> averaging across 50 simulations. The results are tabulated in Table \ref{tab:gaus}, \ref{tab:bin_auc}.

@@

@@

The predictive performances have a consistent pattern across the two distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined. To note, mgcv doesn't support high-dimensional analysis when the number of coefficients are greater than the sample size. mgcv predicts well when $p$ is small or moderate (p = 4, 10), and its performance deteriorates when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are small, moderate, and high. However, when $p$ increase to 200 for the extremely sparse-signal setting, SB-GAM has better performance. Among the proposed methods, BHAM-IWLS performs consistently better than BHAM-CD at the cost of computation time.

### 2021-10-01 Manuscript/04-Real_Data.Rmd
ENAR Submission Version

We apply the proposed models BHAM, fitted with the EM-CD algorithm ,to analyze two real-world metabolomics datasets where the outcomes are binary and continuous respectively. We present the analysis results for the Emory Cardiovascular Biobank data [@Mehta2020], and include the analysis of Weight Loss Maintenance Cohort  data in Appendix A.2. We demonstrate the improved prediction performance compared to the other Bayesian hierarchical additive model, SB-GAM [@Bai2021]. Computation time for cross-validation and optimal model fitting steps are recorded and presented in Table \ref{tab:time_real_data}, where we see the proposed model are more computationally efficient.

```{r results = "asis"}
tbl_real_time %>% 
  xtable(
    caption = "Model fitting time in seconds for two metabolomics data analyses, from Emory Cardiovascular Biobank (ECB) and Weight Loss Maintenance Cohort (WLM). It tabulates the computation time for cross-validation step (CV) and optimal model fitting step (Final), and total computation time (Total) for the proposed model BHAM with EM-CD algorithm (BHAM-CD) and the model of comparison SB-GAM.",
    label= "tab:time_real_data"
  ) %>% 
  print(comment = FALSE, include.rownames = FALSE, caption.placement = "top")

@@

## Emory Cardiovascular Biobank
We use the proposed models BHAM-SSL to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There were initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we performed a univariate screening procedure on the features, via GAM implemented in \texttt{mgcv}, and chose the the top 200 features with smallest p-values. 5-Knot spline additive models for binary outcome are constructed using two different models, the proposed BHAM-SSL and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples were used for prediction performance evaluation, where deviance, area under the ROC curve (AUC), mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ were calculated and tabulated in Table \ref{tab:ECB_res}.

```{r results = "asis"}
rbind(
  'BHAM-SSL' = ECB_bamlasso_cv %>% slice_max(auc) %>% select(-s0),
  'SB-GAM' = ECB_SBGAM_cv %>%  slice_max(auc) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>% 
  xtable(
    caption = "Measures of optimal BHAM-SSL and SB-GAM models for Emory Cardiovascular Biobank by 10-fold cross-validation",
    label = "tab:ECB_res"
  ) %>% 
  print(comment = FALSE, include.rownames = FALSE, caption.placement = "top")

>>>>

We apply the proposed models BHAM, fitted with the EM-CD algorithm ,to analyze two real-world metabolomics datasets where the outcomes are binary and continuous respectively. We present the analysis result for the Emory Cardiovascular Biobank data [@Mehta2020], and include the analysis of Weight Loss Maintenance Cohort  data [@Bihlmeyer2021] in Appendix A.2. We observe a improved prediction performance compared to the other Bayesian hierarchical additive model, SB-GAM [@Bai2021]. Computation time for cross-validation and optimal model fitting steps are recorded and presented in Table \ref{tab:time_real_data}, where we see the proposed model are more computationally efficient.

\begin{table}[ht]
\centering
\caption{Model fitting time in seconds for two metabolomics data analyses, from Emory Cardiovascular Biobank (ECB) and Weight Loss Maintenance Cohort (WLM). It tabulates the computation time for cross-validation step (CV) and optimal model fitting step (Final), and total computation time (Total) for the proposed model BHAM with EM-CD algorithm (BHAM-CD) and the model of comparison SB-GAM.} 
\label{tab:time_real_data}
\begin{tabular}{lcccccc}
  \hline
  & & BHAM-CD & & & SB-GAM &\\
  \cmidrule{2-4}  \cmidrule{5-7}
Data & CV & Final & Total & CV & Final & Total \\ 
  \hline
ECB & 225.23 & 3.02 & 228.25 & 3506.39 & 34.36 & 3540.75 \\ 
  WLM & 1207.98 & 8.03 & 1216.01 & 3116.00 & 32.68 & 3148.68 \\ 
   \hline
\end{tabular}
\end{table}

```{r eval = F, results = "asis"}
tbl_real_time %>% 
  xtable(
    caption = "Model fitting time in seconds for two metabolomics data analyses, from Emory Cardiovascular Biobank (ECB) and Weight Loss Maintenance Cohort (WLM). It tabulates the computation time for cross-validation step (CV) and optimal model fitting step (Final), and total computation time (Total) for the proposed model BHAM with EM-CD algorithm (BHAM-CD) and the model of comparison SB-GAM.",
    label= "tab:time_real_data"
  ) %>% 
  print(comment = FALSE, include.rownames = FALSE, caption.placement = "top")

@@

## Emory Cardiovascular Biobank
We use the proposed models BHAM to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There were initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we performed a univariate screening procedure on the features, via GAM implemented in \texttt{mgcv}, and chose the the top 200 features with smallest p-values. 5-Knot spline additive models for binary outcome are constructed using two different models, the proposed BHAM and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the softwares. Out-of-bag samples were used for prediction performance evaluation, where deviance, area under the ROC curve (AUC), mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ were calculated and tabulated in Table \ref{tab:ECB_res}.

```{r results = "asis"}
rbind(
  'BHAM' = ECB_bamlasso_cv %>% slice_max(auc) %>% select(-s0),
  'SB-GAM' = ECB_SBGAM_cv %>%  slice_max(auc) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>% 
  select(-mae) %>% 
  rename(Deviance = deviance,
         AUC = auc,
         MSE = mse,
         Misclass = misclassification) %>% 
  xtable(
    caption = "Prediction performance of optimal BHAM and SB-GAM models for Emory Cardiovascular Biobank by 10-fold cross-validation, including deviance, area under the curve (AUC), mean squared error (MSE), and misclassification (Misclass)",
    label = "tab:ECB_res"
  ) %>% 
  print(comment = FALSE, include.rownames = FALSE, caption.placement = "top")

### 2021-10-01 Manuscript/05-Discussion.Rmd
ENAR Submission Version

# Discussion
\label{sec:concl}
In the paper, we described a novel generalized additive model using Bayesian hierarchical priors, particularly the proposed spike-and-slab spline prior for bi-level functional selection. Meanwhile, we introduced two optimization based algorithms for model fitting. The algorithms can be easily scale up to address high-dimensional data analysis in a computational efficient manner, but also provide naive inferential supports for many commonly used hierarchical priors, e.g. t-distribution, double exponential distribution, mixture t distribution, and mixture double exponential distribution, in the analyses that the sparse assumption is weak or not necessary. Via simulations, we demonstrated that the proposed model provides as good, if not better, prediction performance compared to some state-of-the-art non-linear modeling devices.

>>>>

# Discussion
\label{sec:concl}
In this article, we describe a novel generalized additive model using Bayesian hierarchical priors, particularly the proposed spike-and-slab spline prior for bi-level functional selection. Meanwhile, we introduce two optimization-based algorithms for model fitting. The algorithms can be easily scale up to address high-dimensional data analysis in a computational efficient manner, but also provide naive inferential supports for many commonly used hierarchical priors, e.g. t-distribution, double exponential distribution, mixture t distribution, and mixture double exponential distribution, in the analyses that the sparse assumption is weak or not necessary. Via simulations, we demonstrated that the proposed model provides as good, if not better, prediction performance compared to some state-of-the-art nonlinear modeling tools.
